---
title: "Practice Worksheet: Gerrymandering"
format: typst
---

Recall the 3 quantitative approaches to measuring gerrymandering discussed in lecture:\
1. Mean - Median Score (Grofman, 1983)\
2. Efficiency Gap (Stephanopolous and McGee, 2015)\
3. Ensembles / Extreme Outliers\
Use the following table to practice applying the first two techniques.\

| District  | D Votes | R Votes | Result |
|:---------:|:-------:|:-------:|:-------|
|     1     |   62    |   38    | D wins |
|     2     |   55    |   45    | D wins |
|     3     |   44    |   56    | R wins |
|     4     |   47    |   53    | R wins |
|     5     |   46    |   54    | R wins |
| **Total** | **254** | **246** |        |

```{r}
#| message: false
#| echo: false
library(tidyverse)
votes <- tibble(District = 1:5,
                d_votes = c(62, 55, 44, 47, 46),
                r_votes = c(38, 45, 56, 53, 54),
                result = c(rep("D wins", 2), rep("R wins", 3)))
```

### Q1. Mean-Median Score

Calculate the mean-median score associated with this particular district breakdown of votes.

```{r}
# add code for Q1 here
dem_votes <- votes$d_votes / (votes$d_votes + votes$r_votes)
dem_mean <- mean(dem_votes)
dem_med <- median(dem_votes)
dem_mm <- dem_mean - dem_med

rep_votes <- votes$r_votes / (votes$r_votes + votes$d_votes)
rep_mean <- mean(rep_votes)
rep_med <- median(rep_votes)
rep_mm <- rep_mean - rep_med

print(paste0("The Democratic mean-median score is ", dem_mm))
print(paste0("The Republican mean-median score is ", rep_mm))
```

### Q2. Efficiency Gap

Calculate the Efficiency Gap (EG) associated with this particular district breakdown of votes.

*Tip*: it may be helpful to write a function that takes a vector of votes across districts for party A and vector of votes for the party B and returns a vector of the number votes wasted by party A across districts.

```{r}
# add code for Q2 here
waste <- function(A, B) { #function takes two vectors, one for each party
  
  total_waste = c() #creates a vector to store the wasted votes for each district
  
  win_marks = c() #creates an vector to store the unique winning threshold for each district
  
  for (i in 1:length(A)) { #iterates through the number of districts
    win_mark = floor((A[i] + B[i]) / 2) + 1 #calculates the winning threshold for each district
    win_marks = c(win_marks, win_mark) #adds to the vector of winning thresholds to have one for each district
  }
    
  for (i in 1:length(A)) { 
    if (A[i] > B[i]) { #if party A vote is greater than party B, this if statement will calculate the wasted vote
    waste = A[i] - win_marks[i] #wasted votes are party A votes - winning threshold
    total_waste = c(total_waste, waste) #add waste votes to the storage vector
    } else { #if party A didn't win
      waste = A[i] #all of the votes are wasted and added as an element to the waste vector
      total_waste = c(total_waste, waste)
    }
  }
  return(total_waste) #returns a vector of waste votes for each district for one party
}

EG <- function(A, B) { #this function takes two party votes again for each district
  wasteA = waste(A, B) #calculates the waste votes in each district for one party
  wasteB = waste(B, A) #calculates the waste votes for the other party
  
  EG = (sum(wasteA) - sum(wasteB)) / (sum(A) + sum(B)) #calculates the EG = total waste votes across party A and B / total votes of A and B
  
  return(EG) #returns the proportion of wasted votes
}
```

```{r}
waste(votes$d_votes, votes$r_votes)
```

```{r}
EG(votes$d_votes, votes$r_votes)
```
